---
title: "Presentation Exercise – Halloween Candy"
format: html
---

# Halloween Candy Power Ranking 
For the presentation exercise, I chose to use the Halloween candy ranking set. 


## Original Graph from "FiveThirtyEight" Halloween Candy Power Ranking 


## Reproduced Graph for Halloween Candy Power Ranking
```{r}

library(tidyverse)
library(here) 

# Load candy dataset using here()
candy <- read_csv(here("presentation-exercise", "candy-data.csv"))

# Quick look at the data
glimpse(candy)

```

```{r}

# Rank candies by win percentage
candy_ranked <- candy %>%
  arrange(desc(winpercent)) %>%
  mutate(competitorname = fct_reorder(competitorname, winpercent))

# Reproduce the horizontal bar chart
ggplot(candy_ranked, aes(x = winpercent, y = competitorname)) +
  geom_col(fill = "#F28E2B") +
  labs(
    title = "The Ultimate Halloween Candy Power Ranking",
    subtitle = "Win percentage based on head-to-head matchups",
    x = "Win Percentage",
    y = NULL
  ) +
  theme_minimal(base_size = 12)

```

 This coding gave me the general type of graph I was looking to recreate, however, it needs a lot of shaping of be publication ready. To adjust the spacing between categories (bars) in my Halloween candy graph, I asked an ChatGPT: "How do I increase or decrease the spacing between categories in a horizontal bar chart in ggplot2?" AI suggested changing the `width` parameter in `geom_col()` and explained that lower `width` values create more space between bars. After refinement in my code I was able to adjust the code this way below. 

```{r}

# Rank candies by percentages with row number
candy_ranked <- candy %>%
  arrange(desc(winpercent)) %>%
  mutate(
    RK = row_number(),
    competitorname = fct_reorder(competitorname, winpercent)
  )

# Horizontal bar chart with text and spacing
ggplot(candy_ranked, aes(x = winpercent, y = competitorname)) +
  geom_col(fill = "#F28E2B", width = 0.7) +  # adjust width for spacing
  geom_text(
    aes(label = round(winpercent, 1)),
    hjust = -0.1,
    size = 3.5
  ) +
  scale_x_continuous(limits = c(0, 90), expand = c(0, 0)) +
  labs(
    title = "The Ultimate Halloween Candy Power Ranking",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )
 
```

Even with this prompt from AI, the recreated graph is having a figure size issue, so from there I tried to adjust the sizing and proportions of this graph and I adjusted the code to this. 

```{r fig.height=16, fig.width=10}
 
 ggplot(candy_ranked, aes(x = winpercent, y = competitorname)) +
  geom_col(fill = "#F28E2B", width = 0.7) +
  geom_text(
    aes(label = round(winpercent, 1)),
    hjust = -0.1,
    size = 3.5
  ) +
  scale_x_continuous(limits = c(0, 90), expand = c(0, 0)) +
  labs(
    title = "The Ultimate Halloween Candy Power Ranking",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )
 
```


## Summary Table: Top 10 Halloween Candies
Moving forward, we want to create publication ready table. For my dataset I presented the top 10 halloween candies based on the dataset.

```{r}
library(dplyr)
library(gt)
library(scales)

# Top 10 candies by win percentage
top10 <- candy %>%
  arrange(desc(winpercent)) %>%
  slice(1:10) %>%
  mutate(
    Rank = row_number()
  ) %>%
  select(Rank, competitorname, winpercent)

# Calculate automatic summary statistic
mean_top10 <- mean(top10$winpercent)

# Create styled table
top10 %>%
  gt() %>%
  tab_header(
    title = md("**Top 10 Halloween Candies by Win Percentage**"),
    subtitle = "Ranked based on head-to-head matchup performance"
  ) %>%
  fmt_number(
    columns = winpercent,
    decimals = 1
  ) %>%
  data_color(
    columns = winpercent,
    colors = col_numeric(
      palette = c("#fde0dd", "#31a354"),
      domain = NULL
    )
  ) %>%
  cols_label(
    competitorname = "Candy",
    winpercent = "Win Percentage"
  ) %>%
  tab_source_note(
    source_note = md(
      paste0(
        "*Mean win percentage (Top 10):* **",
        round(mean_top10, 1),
        "%**"
      )
    )
  ) %>%
  tab_footnote(
    footnote = "Win percentage calculated from 85,000+ head-to-head candy matchups.",
    locations = cells_column_labels(columns = winpercent)
  )
```
