---
title: "Machine Learning Models Exercise"
format: html
editor: visual
---
## Load the data 
```{r}
library(tidyverse)
library(tidymodels)
library(here)
library(GGally)


set.seed(1234)

# Load your cleaned dataset
data <- readRDS(here("ml-models-exercise", "clean_data.rds"))
glimpse(data)
```

```{r}
glimpse(data)        # see the structure
summary(data)        # quick summary of numeric variables
colnames(data)
```

The dataset used here did not contain a RACE variable, so this step was not not applicable.


## Pairwise Correlations
```{r}

# Select continuous variables
cont_vars <- data %>% select(Y, DOSE, AGE, HT, WT)

# Pairwise correlation plot
ggpairs(cont_vars)
```

```{r}
data <- data %>%
  mutate(BMI = 703 * WT / (HT^2))
```

BMI was created combining height and weight, which may better capture body composition than heght or weight alone.


## Preparing the Data for Modeling
```{r}
# Recipe
rec <- recipe(Y ~ ., data = data) %>%
  step_dummy(all_nominal_predictors())  # encode SEX or other factors
```

## Fitting the Models

```{r}
lin_mod <- linear_reg() %>%
  set_engine("lm")

lin_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(lin_mod)

lin_fit <- lin_wf %>% fit(data)
lin_pred <- predict(lin_fit, data) %>% bind_cols(data)
lin_pred
```

```{r}
rmse(lin_pred, truth = Y, estimate = .pred)
```

The linear model provides a baseline for comparison. It assumes a linear relationship between predictors and the outcome. The RMSE from this model is approximately 581, providing a reference point for evaluating more flexible models.


```{r}
lasso_mod <- linear_reg(penalty = 0.1, mixture = 1) %>%
  set_engine("glmnet")

lasso_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(lasso_mod)

lasso_fit <- lasso_wf %>% fit(data)
lasso_pred <- predict(lasso_fit, data) %>% bind_cols(data)
rmse(lasso_pred, truth = Y, estimate = .pred)
```

The LASSO regression produces an RMSE of approximately 582, which very similar RMSE to the linear model. This is expected because the penalty parameter is small, meaning that the model behaves similarly to a standard linear regression.


```{r}
rf_mod <- rand_forest() %>%
  set_engine("ranger", seed = 1234) %>%
  set_mode("regression")

rf_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_mod)

rf_fit <- rf_wf %>% fit(data)
rf_pred <- predict(rf_fit, data) %>% bind_cols(data)
rmse(rf_pred, truth = Y, estimate = .pred)

```

The random forest model produces a much lower RMSE of approximately 347, showing improved performace comapred to the linear and LASSO models as the RMSE is lower. This can be expected as random forests are highly flexible models able to capture complex relationships in data. However, this flexibility can also increase the risk of overfitting with the data.

```{r}
library(ggplot2)

# Function for plotting
plot_pred <- function(df){
  ggplot(df, aes(x = Y, y = .pred)) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    labs(x = "Observed Y", y = "Predicted Y") +
    theme_minimal()
}

plot_pred(lin_pred)
plot_pred(lasso_pred)
plot_pred(rf_pred)
```

The observed versus predicted plots show that the random forest predictions are generally closer to the 45-degree line 
compared to the linear and LASSO models. This indicates better performance within the model as more predictions closely match the observed values.

# Tuning

## LASSO tuning (no cv) 

```{r}
# Recipe (SEX)
model_recipe <- recipe(Y ~ ., data = data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_zv(all_predictors())

# LASSO model (tunable)
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet")

# LASSO workflow
lasso_workflow <- workflow() %>%
  add_model(lasso_spec) %>%
  add_recipe(model_recipe)
```

## Defining your penalty grid
```{r}
penalty_grid <- tibble(
  penalty = 10^seq(-5, 2, length.out = 50)
)
```

```{r}
# Apparent resampling (full dataset)
resamples_app <- apparent(data)

# Tune LASSO over full dataset
lasso_tune_res <- tune_grid(
  lasso_workflow,
  resamples = resamples_app,
  grid = penalty_grid,
  metrics = metric_set(rmse)
)
```

As the LASSO penalty increases, the RMSE increases. This is expected because larger penalties shrink coefficients toward zero, reducing model flexibility and potentially leading to underfitting. With smaller penalties, the model behaves more like a standard linear regression.

However, this tuning is performed on the full dataset without cross-validation, meaning the model is evaluated on the same data used for training. This leads to overly optimistic performance estimates and increases the risk of overfitting.

```{r}
# Random Forest model (tunable)
rf_spec <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 300
) %>%
  set_engine("ranger", seed = 1234) %>%
  set_mode("regression")

# Random Forest workflow 
rf_workflow <- workflow() %>%
  add_model(rf_spec) %>%
  add_recipe(model_recipe)
```

```{r}
# Define 5-fold CV repeated 5 times
cv_folds <- vfold_cv(data, v = 5, repeats = 5)

# Tune LASSO with CV
lasso_cv_res <- tune_grid(
  lasso_workflow,
  resamples = cv_folds,
  grid = penalty_grid,
  metrics = metric_set(rmse)
)

# Random Forest tuning with CV
rf_grid <- grid_regular(
  mtry(range = c(1, 7)),
  min_n(range = c(1, 21)),
  levels = 7
)

rf_cv_res <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(rmse)
)

autoplot(rf_cv_res)
```

When using cross-validation, the RMSE values are generally higher compared to tuning without cross-validation. This is because the model is evaluated on unseen data in each fold, providing a more realistic estimate of performance. Cross-validation helps prevent overfitting and gives a better indication of how the model will perform on new data.

## Select best parameters
```{r}
# Best LASSO penalty
best_lasso <- select_best(lasso_cv_res, metric = "rmse")

# Best RF parameters
best_rf <- select_best(rf_cv_res, metric = "rmse")
```

The best LASSO penalty minimizes RMSE across CV folds. Likewise, the best Random Forest parameters minimize RMSE and will be used to finalize each model .

## Final models 
```{r}
# Final LASSO model
final_lasso <- finalize_workflow(
  lasso_workflow,
  best_lasso
) %>%
  fit(data)

# Final Random Forest model
final_rf <- finalize_workflow(
  rf_workflow,
  best_rf
) %>%
  fit(data)
```

Using the selected optimal parameter, we fit the final LASSO and Random Forest models to the full dataset. These models will be used for predictions and comparison.

## Predictions and RMSE
```{r}
# Predict on full dataset
lasso_preds <- predict(final_lasso, data) %>% bind_cols(data)
rf_preds <- predict(final_rf, data) %>% bind_cols(data)

# RMSE
rmse(lasso_preds, truth = Y, estimate = .pred)
rmse(rf_preds, truth = Y, estimate = .pred)

```

The RMSE values quantify model predition on the full dataset. For this dataset, the LASSO model has an RMSE of ~582, while the Random Forest model has an RMSE of ~467. This indicates that the Random Forest captures patterns in data better, likely due to its ability to model nonlinear relationships that the LASSO model cannot. 

## Observed vs Predicted Plots
```{r}
library(ggplot2)

# LASSO
ggplot(lasso_preds, aes(x = Y, y = .pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Observed vs Predicted (LASSO)", x = "Observed Y", y = "Predicted Y")

# Random Forest
ggplot(rf_preds, aes(x = Y, y = .pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "blue") +
  labs(title = "Observed vs Predicted (Random Forest)", x = "Observed Y", y = "Predicted Y")
```

Observed vs predicted plots visualize model performance. Points are close to the diagonal indicate accurate prediction. LASSO predictions are more linear while Random Forest may capture nonlinear relationships better.




