# Model Fitting Exercise

## Load Data
First we start by loading our data about the drug Mavoglurant.
```{r}
library(tidyverse)
library(tidymodels)
library(here)

#Load our data
drug_data <- read_csv(here("fitting-exercise", "Mavoglurant_A2121_nmpk.csv"))
glimpse(drug_data)
```

The dataset contains 2,678 observations and 17 variables. Each row represents a measurement at a specific time point for a participant. Key variables include ID for participant identification, DV for measured drug concentration, TIME for when the measurement occurred, DOSE for administered dose, and OCC for dosing occasion. Participant characteristics such as AGE, SEX, RACE, WT (weight), and HT (height) are also included. 

## Plot drug concentration over time for each individual
```{r}
ggplot(drug_data, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  geom_line(alpha = 0.3) +
  labs(
    title = "Drug Concentration Over Time",
    x = "Time",
    y = "Drug Concentration (DV)",
    color = "Dose"
  ) +
  theme_minimal()
```

### Facet by Dose
Here we separate each dosing into a separate graph for easier readability with the `facet_wrap` function.

```{r}
ggplot(drug_data, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  facet_wrap(~ DOSE) +
  geom_line(alpha = 0.3) +
  labs(
    title = "Drug Concentration Over Time",
    x = "Time",
    y = "Drug Concentration (DV)",
    color = "Dose"
  ) +
  theme_minimal()
```

This plot shows drug concentration (DV) over time for each participant. Each line represents one individual, and colors indicate different dose levels. Overall, higher doses appear to produce igher drug concentrations over time.

## Keep only the first dosing occasion
The dataset contains an indicator variable OCC representing dosing occasions. Some individuals received the drug more than once (OCC = 1 and OCC = 2). Since the difference between occasions is unclear and we want to avoid complications from repeated dosing, I retain only observations where OCC = 1. This ensures that the data from the participants, is from only one dosing occasion.
```{r}
drug_occ1 <- drug_data %>%
  filter(OCC == 1)

glimpse(drug_occ1)
```

## Remove dosing rows and compute total drug exposure
```{r}
y_data <- drug_occ1 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarize(Y = sum(DV), .groups = "drop")

dim(y_data)
```

To summarize drug exposure for each participant, total drug concentration across was calcualtes acrosss measured time points. This value is labeled as Y, and represents drug concentration for each individual.

```{r}
# Keep the dosing rows, (TIME = 0)
dose_rows <- drug_occ1 %>%
  filter(TIME == 0)

dim(dose_rows)
```

Observations where TIME = 0 correspond to the dosing event for each individual. These rows contain participant characteristics such as dose, age, sex, race, weight, and height. I extract these rows so they can later be merged with the total (calculated) drug exposure variable.

## Merge exposure with participant characteristics
```{r}
# Drug exposure with the participant characteristics
analysis_data <- left_join(y_data, dose_rows, by = "ID")

dim(analysis_data)
```

```{r}
analysis_data <- analysis_data %>%
  mutate(
    SEX = factor(SEX),
    RACE = factor(RACE)
  ) %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)

glimpse(analysis_data)
```

The total exposure variables, labled Y, is merged with participant characteristics by participant ID and create a dataset where each row represents a participant with their demographic characteristic and total drug exposure.

## Summary Statistics 
I begin by examining summary statistics for the cleaned dataset to understand the general distribution of the variables and check for unusual values.
```{r}
summary(analysis_data)
```

# Exploratory Analysis (EDA)
```{r}
# Histogram
ggplot(analysis_data, aes(x = Y)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of Total Drug Concentration")
```

The histogram displays the distribution of total drug exposure (Y). The distribution appears to be skewed right, which is common for pharmacokinetic measurements where some individuals experience higher exposure than others. Most participants exposure values between ~1700 and ~3050.

```{r}
# Scatterplot by dose
ggplot(analysis_data, aes(x = DOSE, y = Y)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Total Drug Exposure vs Dose")
```

This scatterplot looks at the relationship between dose and total drug exposure. Higher doses tend to correspond with higher total exposure, indicating a positive relationship between dose and total drug exposure.

```{r}
# Scatterplot by age 
ggplot(analysis_data, aes(x = AGE, y = Y)) +
  geom_point() +
  theme_minimal()
```

This plot looks whether age is associated with total drug exposure. There does not appear to be a strong relationship between age and total drug exposure in this dataset as the results are extremely varibale.

```{r}
# pairs plots
pairs(analysis_data[, c("Y","DOSE","AGE","WT","HT")])
```

The pairwise plots show that dose appears to have the strongest relationship with total drug exposure (Y),
while other variables such as age, weight, and height show weaker relationships with (Y)

# Model Fitting
Now we will focus on more modeling fitting for this Mavogulrant dataset.

## Linear Model with Y - DOSE
```{r}
# Linear regression with Y as outcome and DOSE as predictor
lm_fit <- lm(Y ~ DOSE, data = analysis_data)
summary(lm_fit)  # shows coefficients and R-squared
```

This linear regression examines the relationship between dose (DOSE) and total drug exposure (Y). The estimated coefficient for dose is 58.2, meaning that for each additional unit of dose, total exposure increases by about 58 units on average, holding everything else constant. The model explains roughly 52% of the variation in exposure (R² = 0.516), and the F-test for the overall model is highly significant (F = 125.6, p < 0.001), indicating that dose is a meaningful predictor of total drug exposure in this dataset. The residuals suggest moderate variability around the fitted values, so while dose is important, other factors may also contribute to differences in exposure.

## Linear Model with Y - All Predictors 
```{r}
# Linear regression with Y as outcome and all predictors
lm_all <- lm(Y ~ DOSE + AGE + SEX + RACE + WT + HT, data = analysis_data)
summary(lm_all)
```

This linear regression model includes dose with additional characteristics (age, sex, race, weight, and height) to predict total drug exposure. The model explain approximately 61.9% of the variation in total drug exposure, higher than the model including only dose. The overal model is staistically significant (F=22.57, p,0.001) which sugest that the predictors altogether help to explain variation in drug exposure.

```{r}
# Compute RMSE and R-squared for the DOSE model
pred_lm <- predict(lm_fit, analysis_data)
rmse_lm <- yardstick::rmse_vec(truth = analysis_data$Y, estimate = pred_lm)
rsq_lm <- yardstick::rsq_vec(truth = analysis_data$Y, estimate = pred_lm)

rmse_lm
rsq_lm

# Y ~ All Predictors
pred_lm_all <- predict(lm_all, analysis_data)
rmse_lm_all <- yardstick::rmse_vec(truth=analysis_data$Y, estimate=pred_lm_all)
rsq_lm_all <- yardstick::rsq_vec(truth=analysis_data$Y, estimate=pred_lm_all)

rmse_lm_all
rsq_lm_all

```

When including all predictors, the RMSE decreases and R-squared increases slightly, suggesting that age, sex, race, weight, and height add some additional explanatory power beyond dose alone.

## Logistic Model with SEX - DOSE 
```{r}
# Logistic regression with SEX as outcome and DOSE as predictor
log_fit <- glm(SEX ~ DOSE, data = analysis_data, family = binomial)
summary(log_fit)
```

This logistic regression model looks at whether drug dose predicts participant sex. The coefficient for dose is not statistically significant (p = 0.192), indicating that dose is not associated with sex in this dataset. Additionally, the small change between the null deviance and residual deviance suggests that dose does not substantially improve the model's ability to predict sex. However, this can be expected as drug dose is not related to participant sex. 

```{r}
# Predicted probabilities
pred_prob <- predict(log_fit, analysis_data, type = "response")

# Convert to predicted classes
pred_class <- ifelse(pred_prob > 0.5, 2, 1)
pred_class <- factor(pred_class, levels = c(1, 2))
sex_factor <- factor(analysis_data$SEX, levels = c(1, 2))

# Accuracy
accuracy_log <- yardstick::accuracy_vec(
  truth = sex_factor,
  estimate = pred_class
)

accuracy_log

# ROC-AUC
roc_auc_log <- yardstick::roc_auc_vec(
  truth = factor(sex_numeric),
  estimate = pred_prob
)
roc_auc_log
```

ROC-AUC evaluates how well the model distinguishes between the two categories of sex. A value close to 0.5 suggests that the model performs little better than random guessing. 


## Logistic Model with SEX - All Predictors
```{r}
# Logistic regression with SEX as outcome and all predictors
log_all <- glm(SEX ~ DOSE + AGE + RACE + WT + HT + Y, data = analysis_data, family = binomial)
summary(log_all)
```

This logistic regression model includes dose, age, race, weight, height, and total drug exposure (Y) to predict participant sex. Most predictors are not statistically significant, indicating that these variables do not strongly explain differences in sex in this dataset. Height is the only predictor that appears statistically significant (p = 0.0027), indicating it may be associated with sex. Overall, many predictors do not contribute to predicting sex.

```{r}
# Predicted probabilities for full logistic model
pred_prob_all <- predict(log_all, analysis_data, type = "response")

# Convert to predicted classes
pred_class_all <- ifelse(pred_prob_all > 0.5, 2, 1)

# Convert to factors with same levels
pred_class_all <- factor(pred_class_all, levels = c(1, 2))
sex_factor <- factor(analysis_data$SEX, levels = c(1, 2))

# Accuracy
accuracy_log_all <- yardstick::accuracy_vec(
  truth = factor(sex_numeric),
  estimate = factor(pred_class_all)
)
accuracy_log_all

roc_auc_log_all <- yardstick::roc_auc_vec(
  truth = sex_factor,
  estimate = pred_prob_all
)
roc_auc_log_all
```

The full logistic model predicts sex with high accuracy (94.2%), meaning most participants are classified correctly. However, the ROC-AUC is very low (0.02), indicating the model performs poorly at distinguishing between males and females. This mismatch occurs because the dataset is heavily imbalanced (many more males than females). Essentially, while the model “guesses” correctly most of the time, it does not truly capture differences between the sexes.