---
title: "Tidy Tuesday: Bird Sightings at Sea"
format: html
editor: visual
---

# Tidy Tuesday Exercise 

```{r}
# Loading the dataset
library(tidyverse)
library(tidymodels)
library(tidytuesdayR)

tuesdata <- tt_load("2026-04-14")
birds <- tuesdata$birds

#overview of the data we have
glimpse(birds)
```

```{r}
summary(birds)
```

```{r}
# EDA top 10 most frequent in the dataset and counts 
birds %>%
  count(species_common_name, sort = TRUE) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(species_common_name, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Species",
    y = "Number of sightings",
    title = "Top 10 Most Frequently Sighted Bird Species"
  )
```

## Research Questions
Based on the explorated data analysis from the data, I will attempt to predict whether a bird sighting has a high or low number of birds observed using bird species and environment/location. This will be a binary classification problem.

```{r}
birds <- birds %>%
  mutate(
    high_count = ifelse(count > median(count, na.rm = TRUE), "high", "low"),
    high_count = as.factor(high_count)
  )
```

The outcome variable was created by converting bird count into a binary classification (high vs low) based on the median value.

```{r}
set.seed(123)

split <- initial_split(birds, prop = 0.7)
train <- training(split)
test <- testing(split)
```

The dataset was split into training (70%) and testing (30%) sets to evaluate model performance on unseen data.

```{r}
rec <- recipe(high_count ~ species_common_name + sex + age +
                n_feeding + n_sitting_on_water + n_flying_past,
              data = train) %>%

  step_string2factor(all_nominal_predictors()) %>%

  step_impute_mode(all_nominal_predictors()) %>%
  step_impute_median(all_numeric_predictors()) %>%

  step_novel(all_nominal_predictors()) %>%
  step_unknown(all_nominal_predictors()) %>%

  step_dummy(all_nominal_predictors()) %>%

  step_zv(all_predictors())
```

A recipe was created for predictors by handling categorical variables.

## Modeling 
### Logistic Regression
```{r}
log_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

log_wf <- workflow() %>%
  add_model(log_model) %>%
  add_recipe(rec)
```

### Random Forest 
```{r}
rf_model <- rand_forest() %>%
  set_engine("ranger") %>%
  set_mode("classification")

rf_wf <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(rec)
```

### Decision Tree
```{r}
tree_model <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

tree_wf <- workflow() %>%
  add_model(tree_model) %>%
  add_recipe(rec)
```

### Cross Validation

```{r}
set.seed(123)

folds <- vfold_cv(train, v = 5)
```

```{r}
log_fit <- fit_resamples(
  log_wf,
  resamples = folds,
  metrics = metric_set(accuracy, roc_auc)
)

rf_fit <- fit_resamples(
  rf_wf,
  resamples = folds,
  metrics = metric_set(accuracy, roc_auc)
)

tree_fit <- fit_resamples(
  tree_wf,
  resamples = folds,
  metrics = metric_set(accuracy, roc_auc)
)
```

```{r}
collect_metrics(log_fit)
collect_metrics(rf_fit)
collect_metrics(tree_fit)
```

Model performance was compared using cross-validated accuracy and ROC AUC. The best-performing model was selected based on a balance of predictive performance and interpretability.