---
title: "Data Exercis Option 2, Synthetic Data"
format: html
---

#  Synthetic Data Analysis of Sleep, Stress, and Academic performance

This dataset was generated by first simulating baseline variables (age, sleep hours, and screen time) from simple distributions. Stress was then constructed as a linear function of these variables, increasing with screen time and decreasing with sleep. GPA was subsequently defined as a function of both sleep (positive association) and stress (negative association), with an added random error term to introduce realistic variability. This structure has known relationships between variables that are then evaluated in the exploratory analysis and regression model.

## Load Packages and Synthetic Dataset

This synthetic dataset looks at the relationship between sleep, stress, screen time, and GPA among 200 college students.
```{r}
library(tidyverse)

set.seed(123)

n <- 200

data <- tibble(
  age = round(rnorm(n, 21, 2)),
  sleep_hours = rnorm(n, 7, 1.2),
  screen_time = rnorm(n, 5, 2)
) %>%
  mutate(
    stress_score = 50 + screen_time * 4 - sleep_hours * 3 + rnorm(n, 0, 5),
    GPA = 3.8 + sleep_hours * 0.12 - stress_score * 0.02 + rnorm(n, 0, 0.2)
  )
```

## Exploratory Analysis
```{r}
# Distribution of GPA
ggplot(data, aes(x = GPA)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Distribution of GPA",
    x = "GPA",
    y = "Count"
  )
```

The distribution of the histogram is approximately normal and symmetrical. Most observations fall within moderate GPA range, with fewer students at lower and higher GPA. 

```{r}
# Sleep and GPA
ggplot(data, aes(x = sleep_hours, y = GPA)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Sleep Hours and GPA",
    x = "Sleep Hours",
    y = "GPA"
  )
```

There is a positive relationship between sleep hours and GPA. Students who report more sleep tend to have higher academic performance. 

```{r}
# Stress and GPA
ggplot(data, aes(x = stress_score, y = GPA)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Stress and GPA",
    x = "Stress Score",
    y = "GPA"
  )
```

There is a negative relationship between stress score and GPA. As stress increases, GPA tends to decrease. 

## Linear Regression Model
```{r}
# Linear Regression Model

model <- lm(GPA ~ sleep_hours + stress_score, data = data)

summary(model)
```

The linear regression model explains a substantial amount of variation in GPA (R² = 0.694). Overall, the model is highly significant (p < 2.2e-16).

Sleep hours has a positive and significant effect on GPA (β = 0.134, p < 2e-16), indicating that more sleep is associated with higher academic performance.

Stress score has a negative and significant effect on GPA (β = -0.021, p < 2e-16), meaning higher stress is associated with lower GPA.
These results match the relationships built into the synthetic dataset.
